Pipit on the Post: Proving Pre- and Post-conditions of Reactive Systems

### 2024-03-30 2024pipit/section/30-Core.tex
good

We generally prove the required properties via abstract transition systems (\autoref{s:transition}).

@@

To ensure that recursive streams have a computational interpretation, we implement a causality restriction, similar to standard Lustre~\cite{caspi1995functional}.
This restriction checks that all recursive streams are guarded by a followed-by delay.
We implement this as a simple syntactic check: each $\xrec{x}{e}$ can only mention $x$ inside a followed-by.
This check is stricter than necessary: for example, the expression $\xrec{x}{(\xlet{x'}{x + 1}{\xfby{0}{x'}})}$ does mention the recursive stream $x$ outside of the delay, but after inlining the let, it would be causal.
We hope to relax this restriction somewhat in future work.

>>>>

We generally prove the required properties by first translating the program to an abstract transition system, as described in \autoref{s:transition}.

@@

To ensure that recursive streams have a computational interpretation, we implement a causality restriction, similar to standard Lustre~\cite{caspi1995functional}.
This restriction checks that all recursive streams are guarded by a followed-by delay.
We implement this as a simple syntactic check: each $\xrec{x}{e}$ can only mention $x$ inside a followed-by.
This check ensures productivity of recursive streams, but can be too strict: for example, the expression $\xrec{x}{(\xlet{x'}{x + 1}{\xfby{0}{x'}})}$ mentions the recursive stream $x$ outside of the delay and is outlawed, but after inlining the let, it would be causal.
We hope to relax this restriction in future work.

### 2024-03-30 2024pipit/section/40-Transition.tex
good

\autoref{f:system-types} shows the types of transition systems.
A transition system is parameterised by its input context and the result type.
It also contains two internal contexts: firstly, the state context describes the private state required to execute the machine; secondly, the free context contains any extra input values that the transition system would like to existentially quantify over.
The free context is used to allow the system to ask for arbitrary values from the environment, when it would not otherwise be able to return a concrete value.

@@

Back to \autoref{f:system-types}, the step-result contains the updated state for the transition system, as well as the result value.
The step-result additionally contains two propositions for the `rely', or assumptions about the execution environment, and `guarantee', or obligations that the transition system must show.
For the transition system corresponding to an expression $e$, these propositions are analogous to the known checked semantics $\semcheck{\Sigma}{\PSValid}{e}$ and unknown checks $\semcheck{\Sigma}{\PSUnknown}{e}$ respectively.

For example, recall again the sum contract:
$$
\sumcontractX{\PSUnknown}{\textit{ints}}
$$

@@

  \begin{tabbing}
  MM \= update: \= = \= update \= \kill
  @let@ sum_def: system (ints: $\ZZ$) $\ZZ$ = \{ \\
  \> state   \> = (sum_fby: $\ZZ$); \\
  \> free  \> = (sum: $\ZZ$); \\
  \> init  \> = \{ sum_fby = 0 \}; \\
  \> step  \> = $\lambda{} i~f~s.$ \{ \\
  \> \> \> update \> = \{ sum_fby = f.sum \}; \\
  \> \> \> value  \> = f.sum; \\
  \> \> \> rely   \> = (f.sum = s.sum_fby + i.ints) $\wedge$ i.ints > 0; \\
  \> \> \> guar   \> = f.sum > 0; \} \}
  \end{tabbing}

The initial state of 0 corresponds to the initial value of the followed-by.
In the step function, argument $i$ refers to the input heap containing i.ints, $f$ refers to the free heap containing the recursive stream f.sum, and $s$ refers to the state heap containing s.sum_fby.
In the rely of the step result, f.sum is constrained to match the body of the recursive stream.
The step result's rely also includes the contract's rely that the input integer is positive.
Finally, the step result's guarantee includes the contract's guarantee that the contract holds.

@@

The translation for contract instantiations is similar, except that the contract body is replaced by an arbitrary value from the free context.
For example, we can use the sum contract to implement the Fibonacci sequence with
$
  @rec@~\fibvar.\, \text{sum}~(\xfby{1}{\fibvar})
$.
This program does not require any input values, so we leave the input context empty.
The state context includes an entry for the $\xfby{1}{\fibvar}$ followed-by expression, but does not include the followed-by expressions inside the contract definition.
Similarly, the free context includes an entry for the recursive stream, and an entry for the contract itself:

  \begin{tabbing}
  MM \= update: \= = \= update \= \kill
  @let@ fib_def: system () $\ZZ$ = \{ \\
  \> state   \> = (fib_fby: $\ZZ$); \\
  \> free  \> = (fib: $\ZZ$; sum_contract: $\ZZ$); \\
  \> init  \> = \{ fib_fby = 1 \}; \\
  \> step  \> = $\lambda{} i~f~s.$ \{ \\
  \> \> \> update \> = \{ fib_fby = f.fib \}; \\
  \> \> \> value  \> = f.fib; \\
  \> \> \> rely   \> = (f.fib = f.sum_contract) \\
  \> \> \>        \> $\wedge$ (s.fib_fby > 0 $\implies$ f.sum_contract > 0); \\
  \> \> \> guar   \> = s.fib_fby > 0; \} \}
  \end{tabbing}

As before, the step result's rely includes the assumption that the recursive stream's value (f.fib) agrees with its result (f.sum_contract).
Additionally, the rely includes the assumption that the contract's rely implies the guarantee: if sum's input (s.fib_fby) is positive, then its output (f.sum_contract) is positive too.
Finally, the step result's guarantee encodes the obligation that the contract's \emph{rely} must hold -- the followed-by's state variable is positive.

@@

We now present the details of the translation.
For causal expressions, the translated transition system is verified to be an abstraction of the original expression's dynamic semantics, and the generated proof obligations correspond to the checked semantics.

@@

\autoref{f:system-translation} defines the translation for expressions.
Values and expressions have no internal state.
For variables, we look for the variable binding in either of the input or free heaps; bindings are unique and cannot occur in both.
We omit the rely and guarantee definitions here; both are trivially true.

@@

To translate a followed-by $\xfby{v}{e}$, we initialise the follow-by's unique binder $x_{@fby@(e)}$ to $v$.
At each step, we return the value in the local state, before updating the local state to the subexpression's new value.

@@

To verify that the translated system satisfies its proof obligations -- that is, the checked properties and contract relies hold --- we can perform induction on the system's sequence of steps.
A system satisfies its proof obligations if, for any sequence of steps that satisfy its rely or assumptions, the system's guarantee also holds for all of the steps.

Inductive proofs on Lustre programs generally use a non-standard definition of induction, as the property we wish to show is a function of the \emph{step result}, rather than being a function of the \emph{state}.
This means that the base case must take a single step from the initial state to be able to state the property that, if the step's rely holds, then its guarantee holds:
\begin{tabbing}
  @MM@\= @MMMM@ \= @MMMMMMMMMMMMM@ \= \kill
  @let@ $\text{inductive_check_base}~(\textit{sys}: \text{system}~\textit{input}~\tau): \text{prop}$ = \\
  \> $\forall (i: \textit{input}) (f: \textit{sys}\text{.free}).$ \\
  \> @let@ $\textit{stp} = \textit{sys}\text{.step}~ i~ f~ \textit{sys}\text{.init}$ @in@ \\
  \> $\textit{stp}\text{.rely} \implies \textit{stp}\text{.guar}$
\end{tabbing}

For the inductive step case, we allow the system to take \emph{two} steps from an arbitrary state, and assume that the first step satisfied the inductive property:
\begin{tabbing}
  @MM@\= @MMMM@ \= @MMMMMMMMMMMMM@ \= \kill
  @let@ $\text{inductive_check_step}~(\textit{sys}: \text{system}~\textit{input}~\tau): \text{prop}$ = \\
  \> $\forall (i_0~ i_1: \textit{input}) (f_0~ f_1: \textit{sys}\text{.free}) (s_0: \textit{sys}\text{.state}).$ \\
  \> @let@ $\textit{stp}_1 = \textit{sys}\text{.step}~ i_0~ f_0~ s_0$ @in@ \\
  \> @let@ $\textit{stp}_2 = \textit{sys}\text{.step}~ i_1~ f_1~ \textit{stp}_1\text{.state}$ @in@ \\
  \> $\textit{stp}_1\text{.rely} \implies \textit{stp}_1\text{.guar} \implies \textit{stp}_2\text{.rely} \implies \textit{stp}_2\text{.guar}$
\end{tabbing}

This inductive scheme also generalises to \emph{k-induction}, which allows the inductive step to assume the previous $k$ steps satisfied the inductive property, rather than just assuming that the one previous step holds.
K-induction is a fairly standard invariant strengthening technique; intuitively, it allows the proof to use more context of the history of execution~\cite{hagen2008scaling,champion2016kind2,gacek2018jkind}.

@@

We prove that the transition system is an abstraction of the dynamic semantics: that is, if the expression evaluates to $v$ under some context, then there exists some execution of the transition system that also results in $v$.
The transition system itself is deterministic, but the free context provides the non-determinism; our theorem  statement existentially quantifies the free heap.

@@

As most expressions do not modify the state heap, the invariant for most expressions simply descends into the subexpressions.
Where new bindings are added, we use the dynamic semantics to extend the context with the new values.
The invariant for follow-by expressions asserts that the initial state of the follow-by is the default value; on subsequent steps, the state corresponds to the dynamic semantics.
With this invariant, we can prove abstraction:

@@

Finally, we can show the main entailment result: that, if the proof obligations hold on the system, then the original program's proof obligations also hold.

@@

The above theorem allows us to \emph{bless} the expression and mark all properties as valid (\autoref{s:core:blessing}).
Importantly, the assumption that the checked properties hold lets us re-use previously-verified properties without re-proving them.

>>>>

\autoref{f:system-types} shows the types of transition systems.
A transition system is parameterised by its input context and the result type.
It also contains two internal contexts: firstly, the state context describes the private state required to execute the machine; secondly, the free context contains any extra input values that the transition system would like to existentially quantify over.
The free context is used to allow the system to ask for arbitrary values from the environment, when it would not otherwise be able to return a concrete value.
% The type $\textit{heap}~c$, for some context c, refers to the type of heaps containing those values.

@@

Back to \autoref{f:system-types}, the step-result contains the updated state for the transition system, as well as the result value.
The step-result additionally contains two propositions; one for the `rely', or assumptions about the execution environment, and another for the `guarantee', or obligations that the transition system must show.
For the transition system corresponding to an expression $e$, these propositions are roughly analogous to the known checked semantics $\semcheck{\Sigma}{\PSValid}{e}$ and unknown checks $\semcheck{\Sigma}{\PSUnknown}{e}$ respectively.

For example, recall again the sum contract:
$$
@let@~\text{sum}~\textit{ints}~=~\sumcontractX{\PSUnknown}{\textit{ints}}
$$

@@

  \begin{tabbing}
  MM \= update: \= = \= update \= \kill
  @let@ sum_def: system (ints: $\ZZ$) $\ZZ$ = \{ \\
  \> state   \> = (sum_fby: $\ZZ$); \\
  \> free  \> = (sum: $\ZZ$); \\
  \> init  \> = \{ sum_fby = 0 \}; \\
  \> step  \> = $\lambda{} i~f~s.$ \{ \\
  \> \> \> update \> = \{ sum_fby = \textit{f}.sum \}; \\
  \> \> \> value  \> = \textit{f}.sum; \\
  \> \> \> rely   \> = (\textit{f}.sum = \textit{s}.sum_fby + \textit{i}.ints) $\wedge$ \textit{i}.ints > 0; \\
  \> \> \> guar   \> = \textit{f}.sum > 0; \} \}
  \end{tabbing}

The initial state of 0 corresponds to the initial value of the followed-by.
In the step function, argument $i$ refers to the input heap containing \textit{i}.ints, $f$ refers to the free heap containing the recursive stream \textit{f}.sum, and $s$ refers to the state heap containing \textit{s}.sum_fby.
In the rely of the step result, \textit{f}.sum is constrained to be the translated body of the recursive stream.
The translated rely also includes the contract's rely that the input integer is positive.
Finally, the translated guarantee includes the contract's guarantee that the output is positive.

@@

The translation for contract instantiations is similar, except that the contract body is replaced by an arbitrary value from the free context.
For example, we can use the sum contract to implement the Fibonacci sequence with
$
  @rec@~\fibvar.\, \text{sum}~(\xfby{1}{\fibvar})
$.
This program does not require any input values, so we leave the input context empty.
The state context includes an entry for the $\xfby{1}{\fibvar}$ followed-by expression, but does not include the followed-by expressions inside the contract definition.
Similarly, the free context includes an entry for the recursive stream, and an entry for the abstract, underspecified value of the contract:

  \begin{tabbing}
  MM \= update: \= = \= update \= \kill
  @let@ fib_def: system () $\ZZ$ = \{ \\
  \> state   \> = (fib_fby: $\ZZ$); \\
  \> free  \> = (fib: $\ZZ$; sum_contract: $\ZZ$); \\
  \> init  \> = \{ fib_fby = 1 \}; \\
  \> step  \> = $\lambda{} i~f~s.$ \{ \\
  \> \> \> update \> = \{ fib_fby = \textit{f}.fib \}; \\
  \> \> \> value  \> = \textit{f}.fib; \\
  \> \> \> rely   \> = (\textit{f}.fib = \textit{f}.sum_contract) \\
  \> \> \>        \> $\wedge$ (\textit{s}.fib_fby > 0 $\implies$ \textit{f}.sum_contract > 0); \\
  \> \> \> guar   \> = \textit{s}.fib_fby > 0; \} \}
  \end{tabbing}

As before, the translated rely includes the assumption that the recursive stream's value (\textit{f}.fib) agrees with its body (\textit{f}.sum_contract).
Additionally, the rely includes the assumption that the contract's rely implies the guarantee: if sum's input (\textit{s}.fib_fby) is positive, then its output (\textit{f}.sum_contract) is positive too.
Finally, the translated guarantee encodes the obligation that the environment satisfies the \emph{contract's rely} -- the input to sum is positive.

@@

We now present the details of the translation.
For causal expressions, the translated transition system is verified to be an abstraction of the original expression's dynamic semantics, and the generated proof obligations imply that the original expression satisfies the checked semantics.

@@

\autoref{f:system-translation} defines the translation for expressions.
Values and variables have no internal state.
For variables, we look for the variable binding in either of the input or free heaps; bindings are unique and cannot occur in both.
We omit the rely and guarantee definitions here; both are trivially true.

@@

To translate a followed-by $\xfby{v}{e}$, we initialise the followed-by's unique binder $x_{@fby@(e)}$ to the followed-by's default value $v$.
At each step, we return the value in the local state \emph{before} updating the local state to the subexpression's new value.

@@

To verify that the translated system satisfies its proof obligations -- that is, the checked properties and contract relies hold --- we can perform induction on the system's sequence of steps.
A system satisfies its proof obligations if, for any sequence of steps that all satisfy its rely or assumptions, the system's guarantee also holds for all of the steps.

Inductive proofs on Lustre programs generally use a non-standard definition of induction, as the property we wish to show is a function of the \emph{step result}, rather than being a function of the \emph{state}.
This means that the base case must take a single step from the initial state to be able to state the property that, if the step result's rely holds, then its guarantee holds:
\begin{tabbing}
  @MM@\= @MMMM@ \= @MMMMMMMMMMMMM@ \= \kill
  @let@ $\text{inductive_check_base}~(\textit{sys}: \text{system}~\textit{input}~\tau): \text{prop}$ = \\
  \> $\forall (i: \text{heap}~\textit{input}) (f: \text{heap}~\textit{sys}\text{.free}).$ \\
  \> @let@ $\textit{stp} = \textit{sys}\text{.step}~ i~ f~ \textit{sys}\text{.init}$ @in@ \\
  \> $\textit{stp}\text{.rely} \implies \textit{stp}\text{.guar}$
\end{tabbing}

For the inductive step case, we allow the system to take \emph{two} steps from an arbitrary state, assuming that both steps satisfy the rely and the first step satisfied the inductive property:
\begin{tabbing}
  @MM@\= @MMMM@ \= @MMMMMMMMMMMMM@ \= \kill
  @let@ $\text{inductive_check_step}~(\textit{sys}: \text{system}~\textit{input}~\tau): \text{prop}$ = \\
  \> $\forall (i_0~ i_1: \text{heap}~\textit{input}) (f_0~ f_1: \text{heap}~\textit{sys}\text{.free}) (s_0: \text{heap}~\textit{sys}\text{.state}).$ \\
  \> @let@ $\textit{stp}_1 = \textit{sys}\text{.step}~ i_0~ f_0~ s_0$ @in@ \\
  \> @let@ $\textit{stp}_2 = \textit{sys}\text{.step}~ i_1~ f_1~ \textit{stp}_1\text{.state}$ @in@ \\
  \> $\textit{stp}_1\text{.rely} \implies \textit{stp}_1\text{.guar} \implies \textit{stp}_2\text{.rely} \implies \textit{stp}_2\text{.guar}$
\end{tabbing}

This inductive scheme also generalises to \emph{k-induction}, which allows the inductive case to assume the previous $k$ steps satisfied the inductive property, rather than just assuming that the one previous step holds.
K-induction is a fairly standard invariant strengthening technique; intuitively, it allows the proof to use more context of the history of execution~\cite{hagen2008scaling,champion2016kind2,gacek2018jkind}.

@@

We prove that the transition system is an abstraction of the dynamic semantics: that is, if the expression evaluates to $v$ under some context, then there exists some execution of the transition system that also results in $v$.
The transition system itself is deterministic, but the free context provides the non-determinism which may occur from underspecified contracts; our theorem statement existentially quantifies the free heap.

@@

As most expressions do not modify the state heap, the invariant for most expressions simply descends into the subexpressions.
Where new bindings are added, we use the dynamic semantics to extend the context with the new values.
The invariant for followed-by expressions asserts that the initial state of the followed-by is the default value; on subsequent steps, the state corresponds to the dynamic semantics.
With this invariant, we can prove abstraction:

@@

Finally, we can show the main entailment result that if the proof obligations hold on the system, then the original program is valid according to the checked semantics:

@@

The above theorem allows us to \emph{bless} the expression and mark all properties as valid (\autoref{s:core:blessing}).
Importantly, the assumption that the checked properties hold lets us re-use previously-verified properties without re-proving them, allowing for modular proofs.

### 2024-03-30 2024pipit/section/50-Extraction.tex
good

Here, the step function takes records describing instances of the input and state contexts, and returns a pair of the result value and the updated state.
The first two bindings correspond to the seeded evaluation with the recursive value for the sum set to $\bot_\ZZ$; as such, the resulting state $s_0$ is invalid.
The last two bindings recompute the state, this time with the correct recursive value.
This duplication of work can often be removed by the partial evaluation and dead-code-elimination which we perform during code extraction.

@@

The generated C code for \emph{sum}\footnote{This interface is for a variant of the sum contract with 32-bit integers instead of unbounded integers.} includes a struct type to hold the state information, as well as reset and step functions; this interface is standard for Lustre compilers~\cite{bourke2017formally,gerard2012modular}:
  \begin{tabbing}
    @struct sum_state { uint32_t sum_fby; }@ \\
    @void   sum_reset(struct sum_state* state);@ \\
    @int    sum_step(struct sum_state* state, uint32_t ints);@
  \end{tabbing}

The reset function takes the pointer to the state struct and sets it to its initial values.
The step function takes the pointer to the state struct and the inputs, and returns the result integer.
The state struct is updated in-place.
The implementations of these functions avoid dynamic (heap) allocation and are suitable for embedded systems.

>>>>

Here, the step function takes heaps of the input and state contexts, and returns a pair of the result value and the updated state.
The first two bindings correspond to the seeded evaluation with the recursive value for the sum set to $\bot_\ZZ$; as such, the resulting state $s_0$ is invalid.
The last two bindings recompute the state, this time with the correct recursive value used in the state.
This duplication of work can often be removed by the partial evaluation and dead-code-elimination which we perform during code extraction.

@@

The generated C code for \emph{sum}\footnote{This interface is for a variant of the sum contract with 32-bit integers instead of unbounded integers.} includes a struct type to hold the state information, as well as reset and step functions:
  \begin{tabbing}
    @struct sum_state { uint32_t sum_fby; }@ \\
    @void   sum_reset(struct sum_state* state);@ \\
    @int    sum_step(struct sum_state* state, uint32_t ints);@
  \end{tabbing}

The reset function takes the pointer to the state struct and sets it to its initial values.
The step function takes the pointer to the state struct and the inputs, and returns the result integer.
The state struct is updated in-place.
The implementations of these functions avoid dynamic (heap) allocation and are suitable for embedded systems.
This interface is standard for Lustre compilers~\cite{bourke2017formally,gerard2012modular} and other synchronous languages.

### 2024-03-30 2024pipit/section/60-Evaluation.tex
good

We were able to encode equivalent properties in Kind2 and in Pipit, aside from some encoding issues.
For example, the specification-only function that finds the next trigger is naturally recursive.
Kind2 does not support recursive functions, but we were able to encode it by introducing a temporary array and using Kind2's array comprehension syntax for scanning over arrays.
Additionally, while the recursive call \emph{increases} the index, the array scan can only depend on values with lower indices.
Below we compare a hypothetical unsupported recursive definition of @next@ (left) with the array-scan encoding (right):

\begin{minipage}{0.4\textwidth}
\begin{verbatim}
next(i) =
  if enabled(triggers,i)
  then i
  else if i >= LENGTH - 1
  then NO_NEXT_TRIGGER
  else next(i + 1);

\end{verbatim}
\end{minipage}
\begin{minipage}{0.55\textwidth}
\begin{verbatim}
next_array[i] =
  if enabled(triggers,LENGTH - 1 - i)
  then LENGTH - 1 - i
  else if i <= 0
  then NO_NEXT_TRIGGER
  else next_array[i - 1];
next = next_array[LENGTH - 1 - index];

\end{verbatim}
\end{minipage}

@@

\autoref{f:evaluation:kind2-runtime} shows the verification runtime for different sizes of arrays; the Pipit version is parametric in the array size, and is thus verified for all sizes of arrays.
We ran these experiments in Docker on an Intel i5-12500 with 32GB of RAM.
Both Kind2 and Pipit developments of the full trigger-fetch logic are roughly the same size, on the order of two-hundred lines of code including comments.
Ignoring whitespace and comments, the Pipit implementation of trigger-fetch has 26 lines of actual executable code, while the Kind2 code has 32.
The majority of the remaining code comprises the definition of valid schedules (34 for Pipit, 28 for Kind2), and the lemma statements and invariants (12 for Pipit, 31 for Kind2), as well as contract statements and boilerplate.

We were able to verify the Kind2 implementation of the complete trigger-fetch mechanism for up to 32 triggers; above that, our verification timed out after one hour.
For reference, the M_TTCAN hardware implementation of TTCAN supports up to 64 triggers~\cite{bosch2019mttcan}.

>>>>

We were able to express equivalent properties in Kind2 and in Pipit, aside from some encoding issues.
For example, the specification-only function that finds the next trigger is naturally recursive.
Kind2 does not support recursive functions, but we were able to encode it by introducing a temporary array and using Kind2's array comprehension syntax for scanning over arrays.
Additionally, while the recursive call \emph{increases} the index, the array scan can only depend on values with lower indices.
\autoref{f:evaluation:recursion} illustrates this encoding with a simplified version of the next-trigger logic.

\begin{figure}
\begin{minipage}{0.45\textwidth}
\begin{tabbing}
  MM \= MM \= \kill
@let rec@ next ($i$: int) ($c$: cycle): \\
  \> Tot \> (option int) \\
  \>     \> (decreases (count - $i$)) = \\
\>  @if@ trigger_enabled $i$~$c$ \\
\>  @then@ Some $i$ \\
\>  @else if@ $i \ge \text{count} - 1$ \\
\>  @then@ None \\
\>  @else@ next $(i + 1)$~$c$
\end{tabbing}
\end{minipage}
\begin{minipage}{0.55\textwidth}
\begin{verbatim}
  function next(index:  int; c: cycle)
       returns (result: int)
  var next_array: int ^ COUNT;
  let
    next_array[i] =
      if trigger_enabled(COUNT - 1 - i, c)
      then COUNT - 1 - i
      else if i <= 0
      then NO_NEXT_TRIGGER
      else next_array[i - 1];
    result =
      next_array[COUNT - 1 - index];
  tel
\end{verbatim}
\end{minipage}
\caption{Left: next-trigger logic in \fstar{}; right: Kind2 encoding as array scan. In \fstar{}, the \emph{Tot $\tau$ (decreases \ldots)} syntax declares a total function with the given termination measure.
In Kind2, the \tt{int\^{}COUNT} syntax declares an array of integers of length @COUNT@, while the @next_array[i]@ declaration defines the elements of the array as a function of the index @i@.
}
\label{f:evaluation:recursion}
\end{figure}

@@

\autoref{f:evaluation:kind2-runtime} shows the verification runtime for different sizes of arrays; the Pipit version is parametric in the array size, and is thus verified for all sizes of arrays.
We ran these experiments in Docker on an Intel i5-12500 with 32GB of RAM.
Both Kind2 and Pipit developments of the trigger-fetch logic are roughly the same size, on the order of two-hundred lines of code including comments.
Ignoring whitespace and comments, the Pipit implementation of trigger-fetch has 26 lines of actual executable code, while the Kind2 code has 32.
The majority of the remaining code comprises the definition of valid schedules (34 for Pipit, 28 for Kind2), and the lemma statements and invariants (12 for Pipit, 31 for Kind2), as well as contract statements and boilerplate.

We were able to verify the Kind2 implementation of the complete trigger-fetch mechanism for up to 32 triggers; above that, our verification timed out after one hour.
For the simplified trigger-fetch mechanism, we were able to verify up to 64 triggers.
For reference, hardware implementations of TTCAN such as M_TTCAN support up to 64 triggers~\cite{bosch2019mttcan}.

### 2024-03-30 2024pipit/section/70-Related.tex
good

Using existing Lustre tools to verify \emph{and} execute the time-triggered CAN driver from \autoref{s:motivation} is nontrivial.
Compiling the triggers array with an unverified compiler such as Lustre~V6~\cite{jahier2016lustre} or Heptagon~\cite{gerard2012modular} is straightforward; however, the verified Lustre compiler Vélus~\cite{bourke2023verified} does not support arrays or a foreign-function interface.
Recent work on translation validation for LustreC~\cite{brun2023equation} also does not yet support arrays.

Verifying the time-triggered CAN driver is trickier, as the restrictions placed on the triggers array --- that triggers are sorted by time-mark, there must be an adequate time-gap between a trigger and its next-enabled, and a trigger's time-mark must be greater-than-or-equal-to its index --- naturally require quantifiers.
As described in \autoref{s:evaluation}, Kind2 does include experimental array and quantifier support, but in our experiments was unable to verify the full logic for arrays up to the 64 triggers, which is the size supported by hardware implementations of TTCAN.
Additionally, due to the limitations that require the constant triggers array to be passed as an argument, compiling the program with Lustre~V6 would result in the entire triggers array being copied to the stack each iteration.

@@

These model-checkers have definite usability advantages over the general-purpose-prover approach offered here: they can often generate concrete counterexamples and implement counterexample-based invariant-generation techniques such as ICE~\cite{garg2014ice} and PDR~\cite{bradley2011sat,een2011efficient}.
However, even when the problem can be expressed, these model-checkers do not provide much assurance that the semantics they use for proofs matches the compiled code.
In the future, we would like to investigate integrating Pipit with a model-checker via an unverified extraction: such an extraction may allow some of the usability benefits such as counterexamples and invariant generation.
If this integration were used solely for debugging and suggesting candidate invariants, then such a change would not expand the trusted computing base.

>>>>

Using existing Lustre tools to verify \emph{and} execute the time-triggered CAN driver from \autoref{s:motivation} is nontrivial.
Compiling the triggers array with an unverified compiler such as Lustre~V6~\cite{jahier2016lustre} or Heptagon~\cite{gerard2012modular} is straightforward; however, the verified Lustre compiler Vélus~\cite{bourke2023verified} does not support arrays, records, or a foreign-function interface.
Recent work on translation validation for LustreC~\cite{brun2023equation} also does not yet support arrays.

Verifying the time-triggered CAN driver is trickier, as the restrictions placed on the triggers array --- that triggers are sorted by time-mark, there must be an adequate time-gap between a trigger and its next-enabled, and a trigger's time-mark must be greater-than-or-equal-to its index --- naturally require quantifiers.
As described in \autoref{s:evaluation}, Kind2 does include experimental array and quantifier support, but in our experiments was unable to verify the full logic for arrays up to the 64 triggers, which is the size supported by hardware implementations of TTCAN.
Additionally, due to the limitations that require the constant triggers array to be passed as an argument, compiling the program with Lustre~V6 would result in the entire triggers array being copied to the stack each iteration, which is unlikely to result in acceptable performance.

@@

These model-checkers have definite usability advantages over the general-purpose-prover approach offered here: they can often generate concrete counterexamples and implement counterexample-based invariant-generation techniques such as ICE~\cite{garg2014ice} and PDR~\cite{bradley2011sat,een2011efficient}.
However, even when the problem can be expressed, these model-checkers do not provide much assurance that the semantics they use for proofs matches the compiled code.
In the future, we would like to investigate integrating Pipit with a model-checker via an unverified extraction: such an extraction may allow some of the usability benefits such as counterexamples and invariant generation.
If this integration were used solely for debugging and suggesting candidate invariants, then such a change would not necessarily expand the trusted computing base --- that is, we could augment our end-to-end verified workflow with \emph{unverified but validated} invariant generation.
